A Graph Theory Based Approach to MDAO Problem Formulation

### 2013-03-19 graph_syntax.tex
a bit more re-org, moved the indegree stuff into the basic graph syntax

\begin{itemize}
    \item Maximal Connectivity
    \item Fundamental Problem Formulation 
    \item Specific Solution Formulation
\end{itemize}

The \emph{maximal connectivity graph} represents the first phase with all the 
analysis codes being considered with all possible connections between them also present. The second graph 
is \emph{fundamental problem formulation graph}, which is the smallest possible graph 
that still fully defines a given problem formulation. Finally, a \emph{specific problem formulation} 
may be represented by including additional edges and node types to represent the 
solution strategy being employed to solve the problem.

The relationship between these three graphs is depicted in Figs.~\ref{f:tree} and \ref{f:hourglass}. 
The tree diagram demonstrates the fact that it is generally possible to obtain 
multiple FPFs from a single maximal connectivity graph. This  may correspond to 
different down--selections of analysis codes, different connections between them, 
or both. Each down--selection shrinks the number of possible FPFs that could be reached 
untill only one is remaining. Then, a single FPF, different SPFs may be obtained by implementing 
different solution strategies. The hourglass shape in Fig. \ref{f:hourglass} illustrates how 
the MCG gets reduced to a single FPF, then multiple possible SPFs exist to solve the problem.
In other words the FPF is obtained from the MCG by removing nodes and edges, 
and the SPF is obtained from the FPF by adding nodes and edges.

\begin{figure}[htb!]
    \centering
    \subfigure[number of possible graphs]{
    \includegraphics[width=2.0in]{images/tree}
    \label{f:tree}
    }
    \subfigure[graph size]{
    \includegraphics[width=2.0in]{images/hourglass}
    \label{f:hourglass}
    }
\caption{The relationship between the MCG, FPF, and SPF.}
\end{figure}

In the following sections we present some graph theory basics, then demonstrate how to build 
the relevant MDAO constructs from the given grah primatives. Next we show how to construct 
the MCG, FPF, and SPF graphs from those constructs and demonstrate methods to transition between 
them.

\subsection{Graph Theory Basics}
The notation used in this work is adapted from Diestel \cite{Diestel2010}. 
A \emph{graph} is a pair $G = (V,E)$ of sets such that $E \subseteq V \times V$, 
which means that the elements of $E$ are 2--element subsets of $V$. The set $V$ 
contains the \emph{vertices} or \emph{nodes} and the set $E$ contains the \emph{edges}.
For a \emph{directed graph*} we construct $E$ as a set of ordered pairs instead 
of a set of sets. Each ordered pair represents an edge starting at the node 
indicated by the first entry and directed to the node indicated by the second 
entry. Edge $e$ = $(x,y)$ may be referred to simply as $xy$ and we say $y = E(x)$.

@@

Let $I$ be a nonempty set such that for each $i \in I$ there is a corresponding set $A_i$. The family $\{A_i \st i \in I\}$ is an indexed family of sets with index $i$ and indexing set $I$, and we can say $\mathcal A = \{A_i \st i \in I\}$\cite{smith2006}. Then the union over this family of sets is
\begin{equation}
\bigcup_{i \in I} A_i = \bigcup_{A \in \mathcal A} A = \{x \st x \in A \txt{ for } A \in \mathcal A\}.
\end{equation}
Lastly, the cardinality of a set $B$ is the number of elements in $B$ and is denoted as $|B|$.

% \subsection{Problem Formulation Concepts to Represent}
% The MDAO problem formulation concepts we wish to represent are:
% \begin{description}
% \item[global input] A global input is a variable that is taken as given and is used by multiple analyses.
% \item[design variable]
% \item[function call] A function call requires a set of inputs to be fulfilled and has certain properties, such as run time, which are independent of the number of outputs actually being used.
% \item[passing of distinct variables] Instead of connecting analyses, the individual variables as are connected
% \item[objective \& constraint] These are variables that must be produced by the problem formulation for it to be valid
% \item[collision] A collision occurs when an input has multiple sources
% \item[hole] A hole occurs when an input has no sources
% \item[coupling] Couple is the mutual dependence between a set of analysis
% \item[multi--fidelity] It may be desirable for the same variable to be calculated by separate analyses and to use both results.
% \end{description}

\section{MDAO Graph Primatives}
There are four node types:  
\begin{description}
\item[variable: ] represents scalar or array data
\item[model:] responsible for mapping inputs to outputs
\item[expression:] simple algebreic combinations of variables (such as for constraint of objective functions)
\item[driver:] constrol structures capable of managing iteration
\end{description}

@@

The driver node and the implicit edge are only allowed to be pressent in SPF graphs. All 
other node and edge types can be present in in any of the three graphs, subject to a 
few restrictions: 
\begin{enumerate}
\item A model node can only have one edge directed to or from another model node.
\item A model node can only have fixed edges directed in or out.
\item A model node must have at least one edge directed in and at least one edge 
  directed out.
\item If a variable node has an outgoing edge to a model node then it may not have 
  any other outgoing edges.
\end{enumerate}

@@

More formally, all inputs that have no incomming fixed or explicit edges are defined
as design variables. When working with a MCG or FPF, only fixed and explicit edges are allowed, 
in which case a design variable is simply a variable node with no incomming edges. For a SPF 
any design variable node would have at least one incomming implicit edge from a driver node.

@@

For node $v \in V$ the edges directed out are given by $E(v)$ and the edges directed into $v$ are given by $E^{-1}(v)$; $E(E(v))$ is denoted as $E^2(v)$, and likewise for additional levels. 
The \emph{indegree} of a node is the number of edges directed in and is denoted as $\txt{deg}^-(v)$, and the \emph{outdegree} is the number of edges directed out and it is denoted as $\txt{deg}^+(v)$.
We also define the \emph{upper indegree limit} 
\begin{equation}
\txt{deg}_u^-(v):V \to \mathbb{N}
\end{equation} 
and the \emph{lower indegree limit}
\begin{equation}
\txt{deg}_l^-(v):V \to \mathbb{N}.
\end{equation}
These user-specified limits set the number of edges that may be directed into a node for a problem formulation to be valid. For example, a variable node $v$ will have $\txt{deg}_u^-(v) = \txt{deg}_l^-(v) = 1$, unless it is a multi--fidelity variable, in which case the user may specify some upper limit higher than one.

To keep track of which nodes and edges are which type, let $T_\txt{node}$ and $T_\txt{edge}$ be sets containing the possible node types and edge types, respectively, and then define mappings $t_\txt{node}:V \to T_\txt{node}$ and $t_\txt{edge}:E \to T_\txt{edge}$ to assign a type to each node and edge.
Then, for example, for a variable node $v$ we have $t_\txt{node}(v) = \txt{`variable'}$.

 % A \emph{path} is a nonempty graph $P = (V,E)$ with $V = \{x_0,x_1,\ldots,x_k\}$ and $E = \{x_0x_1,x_1x_2,\ldots,x_{k-1}x_k\}$, where $x_i$ are all distinct. A graph $G$ is \emph{connected} if any two of its vertices are linked by a path in $G$. A \emph{tree} is a connected and acyclic graph.
%%% this needs to be changed to fit the definition using a directed graph

\section{Building MCG, FPF, and SPF Graphs}

\subsection{Maximal Connectivity Graph}
To construct the maximal connectivity graph, we assume that a set of codes, global inputs, and objectives and constraints (collectively called global outputs). The codes are represented by analysis blocks $A_i=(V_{A_i},E_{A_i}), \ i\in \{1,2,\ldots,m\}=I$, the global inputs are represented a set of variable nodes $V_\txt{in}$, and the global outputs are represented by a set of variable nodes $V_\txt{out}$. We assume that $V_\txt{in}$, $V_\txt{out}$, and $A_i$ are given, and that any potential connection between variables is given in the form of the explicit edges in the set $C_M$. 
Then we may construct the maximal connectivity graph $M=(V_M,E_M)$ as
\begin{IEEEeqnarray*}{rCl}
V_M & = & V_\txt{in} \cup V_\txt{out} \cup \left( \bigcup_{i \in I} V_{A_i} \right), \\
E_M & = & C_M \cup \left( \bigcup_{i \in I} E_{A_i} \right),
\end{IEEEeqnarray*}
The MCG $M$ is uniquely determined by the given set of analysis blocks, the required outputs, and the given global inputs. In the cases where the set of global inputs $I$ is not known a priori, the process of obtaining the FPF will reveal the required inputs, as discussed subsequently.

@@

\subsection{Obtaining the Fundamental Problem Formulation Graph}
In general, there may be multiple different graphs that satisfy the FPF conditions, though there may be none at all. Here, we describe a process for obtaining an FPF by starting with the MCG and disconnecting explicit edges until the FPF conditions are met. Then the problem is reduced to deciding which explicit edges to remove.

%Then, mathematically, the FPF starts with $\mathcal A_1 = \{1,\ldots,m\}$, $I_{F,1} = I$, and $C_{F,1} = C_M$.
Then, mathematically, the FPF starts with $C_{F,0} = C_M$.
First address the nodes where $\txt{deg}^-(v) < \txt{deg}_l^-(v)$ then the ones where $\txt{deg}^-(v) > \txt{deg}_u^-(v)$

@@

    \subsection{Solution Graph Syntax}
    What is the difference between a problem formulation and a problem solution method? Convert from a cyclic graph, to an acyclic graph
    \begin{itemize}
        \item Cycles indicate convergence loops or design variable loops
        \item Problem can't be solved until all loops are *removed* by adding solvers/optimizers
        \item *Special* nodes for solvers and optimizers that *break* loops (from an algorithmic point of view)
        \item FPF represents the minimal amount of information necessary to define a problem
        \item Any solution path grows the graph complexity by adding edges and nodes (or possibly have an empty solution graph, which you build up
        as you remove edges from problem formulation graph?)
    \end{itemize}

>>>>

\begin{itemize}
    \item Maximal Connectivity Graph (MCG)
    \item Fundamental Problem  (FPG)
    \item Problem Solution Graph (SPG)
\end{itemize}

The \emph{maximal connectivity graph} represents the first phase with all 
analysis codes being considered and all possible connections between them also present. The second graph 
is \emph{fundamental problem formulation graph}, which is the smallest possible graph 
that still fully defines a given problem formulation. Finally, a \emph{specific problem formulation} 
may be represented by including additional edges and node types to represent the 
solution strategy being employed to solve the problem.

The relationship between these three graphs is depicted in Figs.~\ref{f:tree} and \ref{f:hourglass}. 
The tree diagram demonstrates the fact that it is generally possible to obtain 
multiple FPGs from a single maximal connectivity graph. This  may correspond to 
different down--selections of analysis codes, different connections between them, 
or both. Each down--selection shrinks the number of possible FPGs that could be reached 
untill only one is remaining. Then, from a single FPG, different PSGs may be obtained by implementing 
different solution strategies. The hourglass shape in Fig. \ref{f:hourglass} illustrates how 
the MCG gets reduced to a single FPG, then multiple possible PSGs exist to solve the problem.
In other words the FPG is obtained from the MCG by removing nodes and edges, 
and the PSG is obtained from the FPG by adding nodes and edges.

\begin{figure}[htb!]
    \centering
    \subfigure[number of possible graphs]{
    \includegraphics[width=2.0in]{images/tree}
    \label{f:tree}
    }
    \subfigure[graph size]{
    \includegraphics[width=2.0in]{images/hourglass}
    \label{f:hourglass}
    }
\caption{The relationship between the MCG, FPG, and PSG.}
\end{figure}

%the relevant MDAO constructs from the given grah primatives. Next we show how to construct 
%the MCG, FPG, and SPF graphs from those constructs and demonstrate methods to transition between 
%them.

\section{Syntax Definition}
In the rest of this section we present the necessary graph theory fundamentals to 
construct any of the different graphs discussed above. 
The notation used in this work is adapted from Diestel \cite{Diestel2010}. 
A \emph{graph} is a pair $G = (V,E)$ of sets such that $E \subseteq V \times V$, 
which means that the elements of $E$ are 2--element subsets of $V$. The set $V$ 
contains the \emph{vertices} or \emph{nodes} and the set $E$ contains the \emph{edges}.
For a \emph{directed graph*} we construct $E$ as a set of ordered pairs instead 
of a set of sets. Each ordered pair represents an edge starting at the node 
indicated by the first entry and directed to the node indicated by the second 
entry. Edge $e$ = $(x,y)$ may be referred to simply as $xy$ and we say $y = E(x)$.

@@

Let $I$ be a nonempty set such that for each $i \in I$ there is a corresponding set $A_i$. 
The family $\{A_i \st i \in I\}$ is an indexed family of sets with index $i$ and 
indexing set $I$, and we can say $\mathcal A = \{A_i \st i \in I\}$\cite{smith2006}. 
Then the union over this family of sets is

\begin{equation}
\bigcup_{i \in I} A_i = \bigcup_{A \in \mathcal A} A = \{x \st x \in A \txt{ for } A \in \mathcal A\}.
\end{equation}
Lastly, the cardinality of a set $B$ is the number of elements in $B$ and is denoted as $|B|$.

There are four node types:  
\begin{description}
\item[variable: ] represents scalar or array data
\item[model:] responsible for mapping inputs to outputs
\item[expression:] simple algebreic combinations of variables (such as for 
  constraint of objective functions)
\item[driver:] constrol structures capable of managing iteration
\end{description}

@@

[[Insert legend for node and edge types]]

The driver node and the implicit edge are only allowed to be pressent in PSGs. All 
other node and edge types can be present in any of the three graphs, subject to a 
few restrictions: 
\begin{enumerate}
\item A model node can only have one edge directed to or from another model node.
\item A model node can only have fixed edges directed in or out.
\item A model node must have at least one edge directed in and at least one edge 
  directed out.
\item If a variable node has an outgoing edge to a model node then it may not have 
  any other outgoing edges.
\item If a variable node has any incomming explicit edges, it may not have any 
  incomming implicit edges. 
\end{enumerate}

For node $v \in V$ the edges directed out are given by $E(v)$ and the edges 
directed into $v$ are given by $E^{-1}(v)$; $E(E(v))$ is denoted as $E^2(v)$, 
and likewise for additional levels. The \emph{indegree} of a node is the number 
of explicit edges directed in and is denoted as $\txt{deg}^-(v)$, and the \emph{outdegree} 
is the number of explicit edges directed out and it is denoted as $\txt{deg}^+(v)$.
We also define the \emph{upper indegree limit}

\begin{equation}
\txt{deg}_u^-(v):V \to \mathbb{N}
\end{equation} 
and the \emph{lower indegree limit}
\begin{equation}
\txt{deg}_l^-(v):V \to \mathbb{N}.
\end{equation}

These user-specified limits govern the number of explicit edges that may be directed into a 
node for a valid graph. Consider a variable node $v$ with
$\txt{deg}_u^-(v) = \txt{deg}_l^-(v) = 1$. In this case, that node must have one, and only one
incomming explicit edge or the graph is not valid. Two possiblities exists for voilating 
the limit conditions: 
\begin{description}
  \item[hole: ] $\txt{deg}_u^-(v) < \txt{deg}_l^-(v)$. The number of incomming edges 
  is less than $\txt{deg}_l^-(v)$. 
  \item[collision: ]$\txt{deg}_u^-(v) > \txt{deg}_u^-(v)$. The number of incomming edges 
  is greater than $ \txt{deg}_u^-(v)$.
\end{description}

\section{Graph Representation of MDAO Features}

@@

More formally, all inputs that have no incomming fixed or explicit edges are defined
as design variables. When working with a MCG or FPG, only fixed and explicit edges are allowed, 
in which case a design variable is simply a variable node with no incomming edges. For a valid PSG 
any design variable node would have at least one incomming implicit edge from a driver node.

@@

\section{Building MCG, FPG, and PSG}

\subsection{Maximal Connectivity Graph}
To construct the maximal connectivity graph, we assume that a set of codes, global inputs, and objectives and constraints (collectively called global outputs). The codes are represented by analysis blocks $A_i=(V_{A_i},E_{A_i}), \ i\in \{1,2,\ldots,m\}=I$, the global inputs are represented a set of variable nodes $V_\txt{in}$, and the global outputs are represented by a set of variable nodes $V_\txt{out}$. We assume that $V_\txt{in}$, $V_\txt{out}$, and $A_i$ are given, and that any potential connection between variables is given in the form of the explicit edges in the set $C_M$. 
Then we may construct the maximal connectivity graph $M=(V_M,E_M)$ as
\begin{IEEEeqnarray*}{rCl}
V_M & = & V_\txt{in} \cup V_\txt{out} \cup \left( \bigcup_{i \in I} V_{A_i} \right), \\
E_M & = & C_M \cup \left( \bigcup_{i \in I} E_{A_i} \right),
\end{IEEEeqnarray*}
The MCG $M$ is uniquely determined by the given set of analysis blocks, the required outputs, and the given global inputs. In the cases where the set of global inputs $I$ is not known a priori, the process of obtaining the FPG will reveal the required inputs, as discussed subsequently.

@@

\subsection{Obtaining the Fundamental Problem Formulation Graph}
In general, there may be multiple different graphs that satisfy the FPG conditions, though there may be none at all. Here, we describe a process for obtaining an FPG by starting with the MCG and disconnecting explicit edges until the FPG conditions are met. Then the problem is reduced to deciding which explicit edges to remove.

%Then, mathematically, the FPG starts with $\mathcal A_1 = \{1,\ldots,m\}$, $I_{F,1} = I$, and $C_{F,1} = C_M$.
Then, mathematically, the FPG starts with $C_{F,0} = C_M$.
First address the nodes where $\txt{deg}^-(v) < \txt{deg}_l^-(v)$ then the ones where $\txt{deg}^-(v) > \txt{deg}_u^-(v)$

@@

    \subsection{Solution Graph Syntax}
    What is the difference between a problem formulation and a problem solution method? Convert from a cyclic graph, to an acyclic graph
    \begin{itemize}
        \item Cycles indicate convergence loops or design variable loops
        \item Problem can't be solved until all loops are *removed* by adding solvers/optimizers
        \item *Special* nodes for solvers and optimizers that *break* loops (from an algorithmic point of view)
        \item FPG represents the minimal amount of information necessary to define a problem
        \item Any solution path grows the graph complexity by adding edges and nodes (or possibly have an empty solution graph, which you build up
        as you remove edges from problem formulation graph?)
    \end{itemize}
